### 2010-03-10 papers/icfp-2010-desc/paper_desc.tex
Desc@ICFP: finish section on Desc.

Our proposal is based on induction-recursion~\cite{dybjer:general-ir,
  dybjer:axiom-ir, dybjer:ir-initial-algebra, dybjer:iir}. However, in
the context of simple data-types, the full power of
induction-recursion is not required. Therefore, we will present a
stripped-down version. As we extend the scope of our universe of
data-types, our presentation will diverge from induction-recursion. We
shall compare both approaches in Section~\ref{sec:discussion}

@@

So far, for conveniency, we have taken as granted the existence of
$\Desc$, as presented in Figure~\ref{fig:desc_universe}. In
particular, we have considered its code, $\DSigma{}{}$, $\DIndx{}$,
and $\DUnit$ as standard type constructors. Hence, we would expect the
type theory to be extended by these constructors and their typing
rules. Although it makes no conceptual difference, the code of
description should rather be read as a \emph{promise}. We promise the
existence of such objects, satisfying the typing rules. It will be the
subject of Section~\ref{sec:desc-levitate} to fulfill these promises.

@@

\begin{structure}
<- Build the fix-point of functors
    <- See examples: need to build their initial algebra
    -> Extend the type theory with Mu/Con [figure]
        <- Straightforward definition of a fixpoint
            <- Interpret D with (Mu D) as subobjects
\end{structure}

@@

\begin{structure}
<- Extending type propagation
    <- Data-type declaration turns into definitions
        -> Straightforward translation to Desc
        -> Creation of a variable referring to the structure
    <- Push Mu to an applied name [figure]
        -> Direct integration into the type propagation machinery
    <- Labelled Mu
        /> Just mention the possibility of labelling, no details required
        -> For the user, objects have names rather than Mu of codes
    -> Coded presentation is practical
        <- The user never see a code
\end{structure}

\begin{structure}
<- Elimination on Mu
    <- We are used to foldD : \forall X. (desc D X -> X) -> mu D -> X
        /> Not dependent
        -> Cannot express some (which one again?) properties
    -> Develop a dependent induction
        <- Everywhere/All
        <- Induction
    -> *Generic*
    ???
\end{structure}

\[
\F{cata} : \PITEL{D}{\Desc}
           \PI{T}{\Set}
           (\descop{D}{T}\To T) \To 
           \Mu{D} \To T 
\]

\[
\begin{array}{lcl}
\F{induction} & : & \PITEL{D}{\Desc}                \\
              &   & \PITEL{P}{\Mu{D} \To \Set}      \\
              &   & \PITEL{m}{\PI{xs}{\descop{D}{(\Mu{D})}} \All{D}{(\Mu{D})}{P}{xs} \To P (\Con{xs})} \\
              &   & \PI{x}{\Mu{D}} P x
\end{array}
\]

>>>>

Our proposal is based on induction-recursion~\cite{dybjer:general-ir,
  dybjer:axiom-ir, dybjer:ir-initial-algebra, dybjer:iir}. However, in
the context of simple data-types, the full power of
induction-recursion is not necessary. Therefore, we will present a
stripped-down version. As we extend the scope of our universe of
data-types, our presentation will diverge from induction-recursion. We
shall compare both approaches in Section~\ref{sec:discussion}

@@

So far, for conveniency, we have taken as granted the existence of
$\Desc$, as presented in Figure~\ref{fig:desc_universe}. In
particular, we have considered its code, $\DSigma{}{}$, $\DIndx{}$,
and $\DUnit$ as type formers, extending the basic type theory. Hence,
in an implementation of this type theory, we would have extended it
with these constructors and their typing rules. Although it makes no
conceptual difference, the code of description should rather be read
as a specification. We promise the existence of such objects,
satisfying the typing rules. It will be the subject of
Section~\ref{sec:desc-levitate} to fulfill this promise, by actually
implementing the specification.

@@

\begin{wstructure}
<- Build the fix-point of functors
    <- See examples: need to build their initial algebra
    -> Extend the type theory with Mu/Con [figure]
        <- Straightforward definition of a fixpoint
            <- Interpret D with (Mu D) as subobjects
\end{wstructure}

%%! Strictly positive types anyone?

So far, we have used our universe of descriptions to build
functors. To illustrate its usage, we have implemented the signature
functors of natural numbers, lists, and binary trees. However, the
class of functors expressible with descriptions enjoy another
property: they all admit a fix-point.

Let us constructively prove this. First of all, for an object $D$
describing a signature functor, we define its fix-point by the type
former $\Mu{D}$. Inhabitants of the fix-point are materialized by the
structure map: $\Con{x}$ inhabits $\Mu{D}$ if $x$ belongs to the
interpretation of $D$ with the carrier of the initial algebra:

@@

For the less categorically inclined reader, we can simply read this
construction as \emph{tying the knot}: the holes left open by the
functor are filled by its own recursive definition. Hence, we have
recovered the data-types we are usually able to define in simply-typed
languages.

\begin{wstructure}
<- Elimination on Mu
    <- We are used to foldD : \forall X. (desc D X -> X) -> mu D -> X
        /> Not dependent
        -> Cannot express some (which one again?) properties
    -> Develop a dependent induction
        <- Everywhere/All
        <- Induction
    -> *Generic*
    ???
\end{wstructure}

Together with the type formers defined above, we would legibly expect
an elimination principle. Following the categorical intuition, we
could be tempted to provide it as a catamorphism:

\[
\F{cata} : \PITEL{D}{\Desc}
           \PI{T}{\Set}
           (\descop{D}{T} \To T) \To 
           \Mu{D} \To T 
\]

Whereas this definition is amply sufficient in the simply-typed world,
it comes short of its promises in the dependently-typed realm. Indeed,
this type is absolutely non dependent. As this operator will be the
corner stone of any inductive definition, we are better off sharping
it before use.

The first step consists in turning $T$ into a predicate
$\Bhab{P}{\Mu{D} \To \Set}$, hence introducing a dependency on the
descriptions. Consequently, we strengthen the algebra $\descop{D}{T}
\To T$ by a similar process. We start with turning the arrow into a
dependenty arrow, hence offering $\Bhab{xs}{\descop{D}{(\Mu{D})}}$ to
further analysis.

Then, we have to translate the notion of algebra in our dependent
setting. This is achieved by the $\All{D}{X}{P}{x}$ predicate that,
intuitively, states that $\Bhab{P}{X \To \Set}$ holds everywhere in
the sub-structures of $x$. Based on that knowledge, we ought to be
able to prove that $P$ holds for $\Con{xs}$. We close our elimination
principle by taking an argument $x$ in $\Mu{D}$ and claiming that the
motive $P$ is realized on $x$.

\[
\begin{array}{lcll}
\F{induction} & : & \multicolumn{2}{l}{\PITEL{D}{\Desc}}                   \\
              &   & \multicolumn{2}{l}{\PITEL{P}{\Mu{D} \To \Set}}         \\
              &   & \PITEL{m}{& \PI{xs}{\descop{D}{(\Mu{D})}}              \\
              &   &           & \All{D}{(\Mu{D})}{P}{xs} \To P (\Con{xs})} \\
              &   & \multicolumn{2}{l}{\PI{x}{\Mu{D}} P x}
\end{array}
\]

At this stage, we are still left with implementing $\F{All}$ and the
$\F{induction}$ principle. For space reason, we refer the reader to
our technical report for the actual code. Having this dependent
catamorphism, we are empowered with the ability to $\emph{compute}$
over descriptions. Note that this induction principle is the first
manifestation of a generic operation over descriptions. Thanks to
$\F{induction}$, any data-type we define automatically comes equipped
with an induction principle.

\begin{wstructure}
<- Extending type propagation
    <- Data-type declaration turns into definitions
        -> Straightforward translation to Desc
        -> Creation of a variable referring to the structure
    <- Labelled Mu
        /> Just mention the possibility of labelling, no details required
        -> For the user, objects have names rather than Mu of codes
    <- Push Mu to an applied name [figure]
        -> Direct integration into the type propagation machinery
    -> Coded presentation is practical
        <- The user never see a code
\end{wstructure}

In this section, we have introduced a universe of description. We have
shown how to manipulate these codes to build data-types. We have also
presented a fix-point construction and its associated dependent
catamorphism. However, one could argue that programming with these
objects is not practical.

Whereas simply-typed languages feature a convenient grammar for
defining data-types and manipulating them, our proposal could be seen
as a step backward. In our setting, the user would have to write a
description instead of a sugared data-type definition. Then, to
recognize an element of $\Nat$ for instance, she would have to
identify $\Mu{}$ of a description as indeed the fix-point of the
functor signature of natural numbers. Finally, she would have to write
codes instead of the natural data-type constructor.

As for finite sets previously, these symptoms clearly indicate that we
suffer from too much type information. Just as for finite, the cure
lies in rationalizing this information. Let us first tackle the issue
of data-type declaration: unsurprisingly, the grammar of data-types we
are used can be straighforwardly translated to the
``sigmas-of-sigmas'' paradigm. Hence, a data-type definition is turned
into the corresponding description. Then, at the level of definitions,
we associate the name of the data-type to the fix-point of the
description. The induction principle as well as other generic
operations follow for free.

Doing so, we seal data-type definition behind a convenient
abstraction. We take advantage of this abstraction barrier to tamper
the second issue. Namely, our user would rather be presented a name
rather that a fix-point of a signature functor. Rather surprisingly,
it is sufficient to \emph{label} the $\Mu{}$ constructor with the
user-provided name. When pretty-printing this object, we just expose
the label, instead of the definition in full.

Regaining the ability to write data-type constructor is simply a
matter of type propagation. We therefore extends the type-checking
framework of Figure~\ref{fig:type-checking} with the following
inference rule:

%% I'm kind of shafted (haha) by this:
%%     - We have not yet introduced tagged descriptions
%%     - Tagged descriptions are much more complex now
%%     - "cases.i" is a syntax I just made up
\[
\Rule{\forall e_i \in \vec{e}, \propag{\push{\descop{cases.i}{\Mu{(\DSigma{(\Enum\: \CN{const})}{\CN{cases}})}}}{e_i}}
                        {e'_i}}
     {\propag{\push{C\: \vec{e}}{\Mu{(\DSigma{(\Enum\: \CN{const})}{\CN{cases}})}}}
             {\Con{(\etag{C}, \widetilde{e'})}}}\;\etag{C} \in \CN{cases}
\]

Hence, in the high-level expression language, the user types a
familiar type constructors, applied to a telescope of arguments. In
the type propagation setting, this expression will be pushed to a
$\Mu{}$ type of a tagged description. It is then straightforward to
lookup the tag and turn the constructor form into the appropriate
code.

This concludes our presentation of the universe of descriptions. In
this section, we have seen how data-types from the simply-typed world
can be expressed in our dependent-type setting. By relying on a
universe construction technique, we benefit from the ability to
inspect and manipulate these codes. This opens some appealing
opportunities for generic programming. Finally, we have demonstrated
that programming with codes is practical. To this end, we have
extended the type propagation machinery. In this setting, we can
define, interact, and build data-types as transparently as in the
simply-typed world.
